---
title: ''
output: pdf_document
---

# Appendix: code

Analyses were conducted using the R Statistical language (version 4.1.2; R Core Team, 2021) on Windows 10 x64 (build 22000). Code is available at _https://github.com/Minh-AnhHuynh/Osteosarcoma-Project_.

Code philosophy follows ["The tidyverse style guide" written by Hadley Wickham. ](https://style.tidyverse.org/) using _styler (v1.7.0)_ to restyle code, and tidyverse environment as much as possible, substituting base R functions to their _dplyr_ and _purrr_ equivalent in order to ensure consistent code readability, usage and naming.

\spacing{1}

```{r tpm, eval=FALSE}
tpm_function <- function(counts, len) {
  x <- counts / len
  return(t(t(x) * 1e6 / colSums(x)))
}
```

```{r}
sessioninfo::session_info("attached")
```


```{r, echo = FALSE, results="asis"}
pkgs <- sessioninfo::package_info("attached", dependencies = FALSE)
df <- tibble::tibble(
  package = pkgs$package,
  version = pkgs$ondiskversion,
  source = gsub("@", "\\\\@", pkgs$source)
)
knitr::kable(df, format = "markdown")
```

```{r echo=FALSE}
print(glue::glue(report::report(sessionInfo())))
```

